---
title: "Introduction to Cognitive Diagnosis Models (CDM)"
date: 2025-02-14
categories: ["Psychometrics", "R", "Statistics"]
---

Cognitive Diagnosis Models (CDMs) are a class of latent variable models used to diagnose the presence or absence of specific fine-grained attributes (skills) required to solve test items.

## Why CDMs?

Unlike traditional IRT models that place students on a continuous scale of ability ($\\theta$), CDMs classify students into latent classes based on mastery profiles.

- **Diagnostic Feedback**: Provides detailed strengths and weaknesses.
- **Remediation**: Helps teachers target specific skills for improvement.

## The DINA Model

The *Deterministic Input, Noisy "And" gate* (DINA) model is one of the most popular CDMs. It assumes that a student must master **all** required attributes to answer an item correctly.

### Key Parameters

1. **Guessing (g)**: Probability of answering correctly despite lacking skills.
2. **Slipping (s)**: Probability of answering incorrectly despite having skills.

## Implementation in R

We can use the `GDINA` package in R to estimate these models.

### 1. Load Data & Package

```{r}
library(GDINA)

# use built-in example data from GDINA
data(sim10GDINA)
dat <- sim10GDINA$simdat
Q <- sim10GDINA$simQ
```

### 2. Fit the Model

```{r}
est <- GDINA(dat = dat, Q = Q, model = "DINA", verbose = FALSE)
summary(est)
```

### 3. Visualize Results

We can plot the item parameters to see which items are the most difficult or prone to slipping.

```{r}
plot(est, what = "IRF", item = 4)
```

> **Note**: This is a simplified example. Real-world applications often require Q-matrix validation and model fit analysis.

In future posts, I will discuss how to handle longitudinal data in CDMs using the methods described in my recent *Language Assessment Quarterly* paper.

